{ \fontfamily{ptm}\selectfont Catalyst Meta Acceleration Framework: The history and the gist of it }

### 2024-11-21 Notebook LaTEX/Catalyst Accelerations/main.tex
>w<, changes

        For a detailed proofs of the estimating sequence with comparison to the accelerated proximal gradient method, see Appendix \ref*{sec:app:exct-acc-ppm}. 

@@

\section{Non-convex extension of Catalyst acceleration}\label{sec:4wd-catalyst}
    The non-convex extension of Catalyst acceleration by Lin 2018 \cite{lin_catalyst_2018} is similar to the convex case in his 2015 paper \cite{lin_universal_2015}. 
    The new algorithm handles function with unknown weak convexity constant $\rho$ using a process called Auto Adapt subroutine. 
    They only claimed convergence to stationary point is claimed for the weakly convex objective. 
    \par 
    Fix $\kappa$ we use the following notations: 
    \begin{align*}
       & \mathcal M (x; y) := F(x) + \frac{\kappa}{2}\Vert x - y\Vert^2 
       \\
       & \mathcal J y := \argmin_x \mathcal M (x; y). 
    \end{align*}
    We define the algorithm and then its convergence claim below. 
    \begin{definition}[Basic 4WD Catalyst Algorithm]\label{def:basic-4wd-catalyst}
        Find any $x_0 \in \text{dom}(F)$. 
        Initialize the algorithm with $\alpha_1 = 1, v_0 = x_0$. 
        For $k \ge 1$, the iterates $(x_k, y_k, v_k)$ are generated by the procedures: 
        \vspace{-0.5em}
        \begin{align*}
            &
            \textcolor{blue}{
                \text{find } \bar x_k \approx 
            }
            \argmin_{x}\left\lbrace
                    \mathcal M(x; \textcolor{violet}{x_{k - 1}})
                \right\rbrace
            \\ &\quad 
            \text{such that: }
                \text{dist}(\mathbf 0, \partial \mathcal M(\bar x_k; \textcolor{violet}{x_{k - 1}})) 
                \le 
                \kappa\Vert \textcolor{blue}{\bar x_k}- \textcolor{violet}{x_{k - 1}}\Vert, 
                \mathcal M(\textcolor{blue}{\bar x_k}; \textcolor{violet}{x_{k - 1}}) 
                \le F(\textcolor{violet}{x_{k - 1}}); 
            \\
            & y_k = \alpha_k v_{k - 1} + (1 - \alpha_k) \textcolor{violet}{x_{k - 1}};
            \\
            & 
            \textcolor{red}{\text{find }\tilde x_k \approx}
            \argmin_{x} \left\lbrace
                \mathcal M(x; y_k) 
            \right\rbrace
            \text{ such that: }
                \dist\left(
                    \mathbf 0, \partial \mathcal M(\textcolor{red}{\tilde x_k}; y_k)
                \right) 
                \le \frac{\kappa}{k + 1}\Vert \textcolor{red}{\tilde x_k} - y_k\Vert
            ;
            \\
            & v_{k} = \textcolor{violet}{x_{k - 1}} + \frac{1}{\alpha_k}(
                    \textcolor{red}{\tilde x_k} - \textcolor{violet}{x_{k - 1}}
                );
            \\
            & 
            \text{find } \alpha_{k + 1} \in (0, 1): 
            \frac{1 - \alpha_{k + 1}}{\alpha_{k + 1}^2} = \frac{1}{\alpha_k^2};
            \\
            & \text{choose } \textcolor{violet}{x_k} \text{ such that:  } 
            f(\textcolor{violet}{x_k}) = \min(f(\textcolor{blue}{\bar x_k}), f(\textcolor{red}{\tilde x_k})). 
        \end{align*}
    \end{definition}

    \begin{theorem}[Basic 4WD Catalyst Convergence]\label{thm:basic-4wd-catalyst}
        Let $(x_k, v_k, y_k)$ be generated by the basic Catalyst algorithm. 
        If $F$ is $\kappa$ weakly convex and bounded below, then $x_k$ converges to a stationary point where
        \begin{align*}
            \min_{j = 1, \cdots, N} \dist^2(\mathbf 0, \partial F(\bar x_j))
            \le \frac{8 \kappa}{N}(F(x_0) - F^*). 
        \end{align*}
        And when $F$ is convex, $F(x_k) - F^*$ converges at a rate of $\mathcal O(k^{-2})$. 
    \end{theorem}
    \begin{remark}

@@

\appendix
\section{Appendix}

\subsection{Theorems and claims for accelerated proximal gradient}\label{app:sec:thm-claim-acc-prox-grad}
    Throughout this section, $F = g + f$ is an additive composite objective function with $g$ convex, $f$ $L$-lipschitz smooth and $\mu \ge0$ strongly convex. 
    The notations here are 
    \begin{align*}
        \mathcal M^{L^{-1}}(x; y) 
        &:= 
        F(x) + \frac{L}{2}\Vert x - y\Vert^2
        \\
        \widetilde{ \mathcal M}^{L^{-1}}(x; y) 
        &:= g(x) + f(y) 
        + 
        \left\langle \nabla f(x), x - y\right\rangle 
        + 
        \frac{L}{2}\Vert x - y\Vert^2
        \\
        \widetilde{\mathcal J}_{L^{-1}}y 
        &:= \argmin_{x} \widetilde{\mathcal M}^{L^{-1}}(x; y)
        \\
        \widetilde{\mathcal G}_{L^{-1}}(y)
        &:= L\left(I - \widetilde{\mathcal J}_{L^{-1}}\right)y. 
    \end{align*}

@@

\section{Proofs for 4WD Catalyst Acceleration}
    In this section we adopt the same notations as in 
    \hyperref[sec:4wd-catalyst]{Section \ref*{sec:4wd-catalyst}}. 
    We prove
    \hyperref[thm:basic-4wd-catalyst]{Theorem \ref*{thm:basic-4wd-catalyst}}
    in this section. 
    The lemma below characterize the lower and upper bound on the $(\alpha_k)_{k \ge 0}$ which ultimately control the convergence rate. 
    \begin{lemma}[Bounds of the inverted FISTA sequence]\label{app:lemma:momentum-sequence-bounds}
        With base case $\alpha_1 = 1$
        If the sequence $\alpha_k$ has for all $k\ge 1$: 
        \begin{align*}
            \alpha_{k + 1} = \frac{\sqrt{\alpha_k^4 + 4\alpha_k^2} - \alpha_k^2}{2}, \alpha_1 = 1
        \end{align*}
        then for all $k \ge0$: 
        \begin{align*}
            \frac{\sqrt{2}}{k + 1} \le \alpha_k \le \frac{2}{k + 1}. 
        \end{align*}
    \end{lemma}
    \begin{proof}
        The proof of the upper bound is by inductive hypothesis, assume $\alpha_k \ge 2/(k + 1)$ then consider
        \begin{align*}
            \alpha_{k + 1} &= 
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} - \alpha_k^2}{2}
            = 
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} - \alpha_k^2}{2}
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} + \alpha_k^2}{\sqrt{\alpha_k^4 + 4 \alpha_k^2} + \alpha_k^2}
            \\
            &= \frac{\alpha_k^4 + 4\alpha_k^2 - \alpha_k^4}{2(\sqrt{\alpha_k^4 + 4 \alpha_k^2)} + \alpha_k^2)}
            \\
            &= \frac{2}{\sqrt{1 + 4 \alpha_k^{-2}} + 1} 
            \le \frac{2}{\sqrt{1 + (k + 1)^2} + 1} \le \frac{2}{k + 2}. 
        \end{align*}
        The lower bound is given by the hidden recursive definition of $\alpha_k$: 
        \begin{align*}
            \alpha_{k + 1}^2 &= (1 - \alpha_{k + 1})\alpha_k^2 = \alpha_1^2
            \\
            &\ge \alpha_1^2 \prod_{i = 2}^{k + 1}(1 - \alpha_i)
            \\
            &\ge \prod_{i = 2}^{k + 1}\left(
                1 - \frac{2}{i + 1}
            \right) = \prod_{i = 2}^{k + 1} \left(\frac{i - 1}{i + 1}\right)
            \\
            &= \frac{2}{(k + 1)(k + 2)}  \ge \frac{2}{(k + 2)^2}. 
        \end{align*}
    \end{proof}
    \begin{remark}
        $a_k^{-1}$ would just be the FISTA sequence. 
    \end{remark}

@@

    \end{proof}
    \subsection{Convergence proof of the basic 4WD-Catalyst}
        This subsection proves 
        \hyperref[thm:basic-4wd-catalyst]{Theorem \ref*{thm:basic-4wd-catalyst}}
        which is about the convergence of 
        \hyperref[def:basic-4wd-catalyst]{Definition \ref*{def:basic-4wd-catalyst}}.

@@

            One may use KL-Convergence frameworks to strengthen the convergence claim of the Catalyst by establishing convergence to a limiting stationary point. 

>>>>

        For a detailed proof of the estimating sequence with comparison to the accelerated proximal gradient method, see Appendix \ref*{sec:app:exct-acc-ppm}. 

@@

\section{Non-convex extension of Catalyst acceleration}\label{sec:4wd-catalyst}
    The non-convex extension of Catalyst acceleration by Lin 2018 \cite{lin_catalyst_2018} is similar to the convex case in his 2015 paper \cite{lin_universal_2015}. 
    The new algorithm handles function with unknown weak convexity constant $\rho$ using a process called Auto Adapt subroutine. 
    They only claimed convergence to stationarity for a weakly convex objective. 
    \par 
    Fix $\kappa$ we use the following notations: 
    \begin{align*}
       & \mathcal M (x; y) := F(x) + \frac{\kappa}{2}\Vert x - y\Vert^2 
       \\
       & \mathcal J y := \argmin_x \mathcal M (x; y). 
    \end{align*}
    We define the algorithm and then its convergence claim below. 
    \begin{definition}[Basic 4WD Catalyst Algorithm]\label{def:basic-4wd-catalyst}
        Find any $x_0 \in \text{dom}(F)$. 
        Initialize the algorithm with $\alpha_1 = 1, v_0 = x_0$. 
        For $k \ge 1$, the iterates $(x_k, y_k, v_k)$ are generated by the procedures: 
        \vspace{-0.5em}
        \begin{align*}
            &
            \textcolor{blue}{
                \text{find } \bar x_k \approx 
            }
            \argmin_{x}\left\lbrace
                    \mathcal M(x; \textcolor{violet}{x_{k - 1}})
                \right\rbrace
            \\ &\quad 
            \text{such that: }
                \text{dist}(\mathbf 0, \partial \mathcal M(\bar x_k; \textcolor{violet}{x_{k - 1}})) 
                \le 
                \kappa\Vert \textcolor{blue}{\bar x_k}- \textcolor{violet}{x_{k - 1}}\Vert, 
                \mathcal M(\textcolor{blue}{\bar x_k}; \textcolor{violet}{x_{k - 1}}) 
                \le F(\textcolor{violet}{x_{k - 1}}); 
            \\
            & y_k = \alpha_k v_{k - 1} + (1 - \alpha_k) \textcolor{violet}{x_{k - 1}};
            \\
            & 
            \textcolor{red}{\text{find }\tilde x_k \approx}
            \argmin_{x} \left\lbrace
                \mathcal M(x; y_k) 
            \right\rbrace
            \text{ such that: }
                \dist\left(
                    \mathbf 0, \partial \mathcal M(\textcolor{red}{\tilde x_k}; y_k)
                \right) 
                \le \frac{\kappa}{k + 1}\Vert \textcolor{red}{\tilde x_k} - y_k\Vert
            ;
            \\
            & v_{k} = \textcolor{violet}{x_{k - 1}} + \frac{1}{\alpha_k}(
                    \textcolor{red}{\tilde x_k} - \textcolor{violet}{x_{k - 1}}
                );
            \\
            & 
            \text{find } \alpha_{k + 1} \in (0, 1): 
            \frac{1 - \alpha_{k + 1}}{\alpha_{k + 1}^2} = \frac{1}{\alpha_k^2};
            \\
            & \text{choose } \textcolor{violet}{x_k} \text{ such that:  } 
            f(\textcolor{violet}{x_k}) = \min(f(\textcolor{blue}{\bar x_k}), f(\textcolor{red}{\tilde x_k})). 
        \end{align*}
    \end{definition}

    \begin{theorem}[Basic 4WD Catalyst Convergence]\label{thm:basic-4wd-catalyst}
        Let $(x_k, v_k, y_k)$ be generated by the basic Catalyst algorithm. 
        If $F$ is $\kappa$ weakly convex and bounded below, then $x_k$ converges to stationarity where
        \begin{align*}
            \min_{j = 1, \cdots, N} \dist^2(\mathbf 0, \partial F(\bar x_j))
            \le \frac{8 \kappa}{N}(F(x_0) - F^*). 
        \end{align*}
        And when $F$ is convex, $F(x_k) - F^*$ converges at a rate of $\mathcal O(k^{-2})$. 
    \end{theorem}
    \begin{remark}
        Convergence to stationarity is strictly weaker than convergence to any stationarity point of $F$. 
        Read 
        \hyperref[app:sec:ncnvx-catalyst]{Section \ref*{app:sec:ncnvx-catalyst}} for more information on proof of this claim. 
    \end{remark}

@@

\appendix
\section{Theorems and claims for accelerated proximal gradient}\label{app:sec:thm-claim-acc-prox-grad}
    Throughout this section, $F = g + f$ is an additive composite objective function with $g$ convex, $f$ $L$-lipschitz smooth and $\mu \ge0$ strongly convex. 
    The notations here are 
    \begin{align*}
        \mathcal M^{L^{-1}}(x; y) 
        &:= 
        F(x) + \frac{L}{2}\Vert x - y\Vert^2
        \\
        \widetilde{ \mathcal M}^{L^{-1}}(x; y) 
        &:= g(x) + f(y) 
        + 
        \left\langle \nabla f(x), x - y\right\rangle 
        + 
        \frac{L}{2}\Vert x - y\Vert^2
        \\
        \widetilde{\mathcal J}_{L^{-1}}y 
        &:= \argmin_{x} \widetilde{\mathcal M}^{L^{-1}}(x; y)
        \\
        \widetilde{\mathcal G}_{L^{-1}}(y)
        &:= L\left(I - \widetilde{\mathcal J}_{L^{-1}}\right)y. 
    \end{align*}

@@

\section{Proofs for 4WD Catalyst Acceleration}\label{app:sec:ncnvx-catalyst}
    In this section we adopt the same notations as in 
    \hyperref[sec:4wd-catalyst]{Section \ref*{sec:4wd-catalyst}}. 
    We prove
    \hyperref[thm:basic-4wd-catalyst]{Theorem \ref*{thm:basic-4wd-catalyst}}
    in this section. 
    The lemma below characterize the lower and upper bound on the $(\alpha_k)_{k \ge 0}$ which ultimately control the convergence rate. 
    \begin{lemma}[Bounds of the inverted FISTA sequence]\label{app:lemma:momentum-sequence-bounds}
        With base case $\alpha_1 = 1$
        If the sequence $\alpha_k$ has for all $k\ge 1$: 
        \begin{align*}
            \alpha_{k + 1} = \frac{\sqrt{\alpha_k^4 + 4\alpha_k^2} - \alpha_k^2}{2}, \alpha_1 = 1
        \end{align*}
        then for all $k \ge0$: 
        \begin{align*}
            \frac{\sqrt{2}}{k + 1} \le \alpha_k \le \frac{2}{k + 1}. 
        \end{align*}
    \end{lemma}
    \begin{proof}
        The proof of the upper bound is by inductive hypothesis, assume $\alpha_k \ge 2/(k + 1)$ then consider
        \begin{align*}
            \alpha_{k + 1} &= 
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} - \alpha_k^2}{2}
            = 
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} - \alpha_k^2}{2}
            \frac{\sqrt{\alpha_k^4 + 4 \alpha_k^2} + \alpha_k^2}{\sqrt{\alpha_k^4 + 4 \alpha_k^2} + \alpha_k^2}
            \\
            &= \frac{\alpha_k^4 + 4\alpha_k^2 - \alpha_k^4}{2(\sqrt{\alpha_k^4 + 4 \alpha_k^2)} + \alpha_k^2)}
            \\
            &= \frac{2}{\sqrt{1 + 4 \alpha_k^{-2}} + 1} 
            \le \frac{2}{\sqrt{1 + (k + 1)^2} + 1} \le \frac{2}{k + 2}. 
        \end{align*}
        The lower bound is given by the hidden recursive definition of $\alpha_k$: 
        \begin{align*}
            \alpha_{k + 1}^2 &= (1 - \alpha_{k + 1})\alpha_k^2 = \alpha_1^2
            \\
            &\ge \alpha_1^2 \prod_{i = 2}^{k + 1}(1 - \alpha_i)
            \\
            &\ge \prod_{i = 2}^{k + 1}\left(
                1 - \frac{2}{i + 1}
            \right) = \prod_{i = 2}^{k + 1} \left(\frac{i - 1}{i + 1}\right)
            \\
            &= \frac{2}{(k + 1)(k + 2)}  \ge \frac{2}{(k + 2)^2}. 
        \end{align*}
    \end{proof}
    \begin{remark}
        $a_k^{-1}$ would just be the FISTA sequence. 
    \end{remark}

@@

    \subsection{Convergence proof of the basic 4WD-Catalyst}
        This subsection proves 
        \hyperref[thm:basic-4wd-catalyst]{Theorem \ref*{thm:basic-4wd-catalyst}}
        which is about the convergence of 
        \hyperref[def:basic-4wd-catalyst]{Definition \ref*{def:basic-4wd-catalyst}}.

@@

            It's tempting to think whether KL conditions assists with convergence to a stationary point for Nesterov's accelerated gradient method. 
            Surveys conducted by the writer in the literatures showed that this remains a huge mystery in general. 
